Spin-orbit-coupling-induced squeezing in the optical lattice clock

### 2018-07-26 soc_squeezing/soc_squeezing_notes.tex
modified some figure parameters and added 'bare' lattice Hamiltonian prior to Hubbard model approximation

We start with two-level atoms loaded into the ground band of an
optical lattice, and consider interrogating these atoms by a linearly
polarized plane-wave laser tuned to the relevant electronic (atomic)
transition, with wavenumber projection $\phi$ onto the lattice axis.
After performing a gauge transformation which shifts the on-axis
momenta of ground (excited) states by $\phi/2$ ($-\phi/2$), and
eliminating (i.e.~turning off) the interrogation laser, one can arrive
at the Hamiltonian
\begin{align}
  H_{\t{SOC}}
  = -2J_0\sum_{q,s} \cos\p{q+s\phi/2} \hat c_{qs}^\dag \hat c_{qs}
  \label{eq:H_SOC_start}
\end{align}
where $q$ indexes quasi-momentum along the lattice axis,
$s\in\set{\g\leftrightarrow-1,\e\leftrightarrow1}$ labels the
electronic state, $J_0$ is the (positive) ground-band tunneling rate,
$c_{q\sigma}$ is a fermionic annihilation operator, and we work in
units with the lattice spacing $a=1$.  Defining
\begin{align}
  \1_q \equiv \hat c_{q,\g}^\dag \hat c_{q,\g}
  + \hat c_{q,\e}^\dag \hat c_{q,\e},
  &&
  \sigma_q^j \equiv \sum_{\alpha,\beta}
  \hat c_{q\alpha}^\dag \sigma^j_{\alpha\beta} \hat c_{q\beta},
  \label{eq:pseudospin}
\end{align}
for Pauli matrices $\sigma^j$ (i.e.~such that $\sigma_q^j$ is a
pseudo-spin-1/2 Pauli operator on the electronic degrees of freedom)
and expanding the cosine in \eqref{eq:H_SOC_start}, we thus find
\begin{align}
  H_{\t{SOC}}
  = -\sum_q \p{\epsilon_q \1_q + \f12 h_q \sigma_q^z},
  &&
  \epsilon_q \equiv 2J_0 \cos\p{\phi/2} \cos q,
  &&
  h_q \equiv -4J_0 \sin\p{\phi/2} \sin q.
\end{align}
When all dynamics conserve the occupations of all quasi-momenta, the
first term in this Hamiltonian merely contributes a constant energy
shift which we can safely neglect, leaving us with the effective
inhomogeneous-field Hamiltonian
\begin{align}
  H_{\t{SOC}} = -\sum_q h_q s_q^z,
  &&
  s_q^z \equiv \f12 \sigma_q^z
  \label{eq:H_SOC}
\end{align}

@@

At ultracold temperatures inter-atomic interactions are dominated by
$s$-wave collisions, which are captured by the Hamiltonian
\begin{align}
  H_{\t{int}} = G \int \d^3x~
  \hat\psi_\e^\dag\p{x} \hat\psi_\g^\dag\p{x}
  \hat\psi_\g\p{x} \hat\psi_\e\p{x},
  &&
  G \equiv \f{4\pi a_{\e\g^-}}{m_A},
\end{align}
where $\hat\psi_\sigma$ is a fermionic field operator for atoms in
electronic state $\sigma\in\set{\g,\e}$, $a_{eg}^-$ is an scattering
length, and $m_A$ is the mass of a single atom.  Considering only
occupation of the ground band in a lattice, we can expand the field
operators as
\begin{align}
  \hat\psi_\sigma\p{x} = \sum_q \phi_q\p{x} \hat c_{q\sigma},
\end{align}
for momentum-space (Bloch) wavefunctions $\phi_q$ and fermionic
annihilation operators $\hat c_{q\sigma}$.  The interaction
Hamiltonian then becomes
\begin{align}
  H_{\t{int}} = G \sum_{p,q,r,s} K^{pq}_{rs}
  \hat c_{r,\e}^\dag \hat c_{s,\g}^\dag \hat c_{q,\g} \hat c_{p,\e},
  &&
  K^{pq}_{rs} \equiv \int \d^3x~
  \phi_r\p{x}^* \phi_s\p{x}^* \phi_q\p{x} \phi_p\p{x}.
\end{align}
In a lattice with $L$ sites centered on positions $x_j$ (i.e.~for
$j=1,\cdots L$ indexing lattice site) and localized (Wannier)
wavefunctions $w_j$, we can expand
$\phi_p\p{x}=L^{-1/2}\sum_je^{-ipx_j}w_j\p{x}$, which implies
\begin{align}
  K^{pq}_{rs} = L^{-2} \sum_{j,k,\ell,m} \int \d^3x~
  \exp\sp{-i\p{px_j+qx_k-rx_\ell-sx_m}}
  w_\ell\p{x}^* w_m\p{x}^* w_k\p{x} w_j\p{x}.
\end{align}
If we now assume that we can neglect integrals without $j=k=\ell=m$,
which is equivalent to neglecting inter-site interactions and
interaction-assisted hopping, then we have
\begin{align}
  K^{pq}_{rs} \approx L^{-2} \sum_j \int \d^3x~
  \exp\sp{-i\p{p+q-r-s}x_j} \abs{w_j\p{x}}^4,
\end{align}
where a stationary phase approximation now forces $p+q-r-s=0$ (mod
$2\pi$ in units with the lattice spacing $a=1$), which is equivalent
to conservation of total momentum.  In terms of the Kronecker delta
$\delta_{p+q,r+s}$ enforcing $p+q=r+s$, it follows that
\begin{align}
  K^{pq}_{rs} \approx \delta_{p+q,r+s} L^{-2}
  \sum_j \int \d^3x~ \abs{w_j\p{x}}^4
  = \delta_{p+q,r+s} L^{-1} \int \d^3x~ \abs{w_0\p{x}}^4,
\end{align}
which motivates the definitions
\begin{align}
  U \equiv G \int \d^3x~ \abs{w_0\p{x}}^4,
\end{align}
in order to express the interaction Hamiltonian in the simple form
\begin{align}
  H_{\t{int}} = \f{U}{L} \sum_{p,q,r,s}
  \delta_{p+q,r+s} \hat c_{r,\e}^\dag \hat c_{s,\g}^\dag
  \hat c_{q,\g} \hat c_{p,\e}.
  \label{eq:H_int_full}
\end{align}

@@

If the single-particle energy spacings are large compared to the
strength of inter-particle interactions, or equivalently if
$J_0\gtrsim U$, then by the secular approximation we can neglect terms
in $H_{\t{int}}$ which do not conserve the sum of single-particle
energies.  By solving this single-particle energy conservation
condition, one finds that we must have $\set{p,q}\ne\set{r,s}$; that
is, atoms in different clock states can only interact via
\begin{enumerate*}[label=(\roman*)]
\item direct density-density terms, and
\item terms which exchange their momenta.
\end{enumerate*}
The surviving terms in \eqref{eq:H_int_full} are thus
\begin{align}
  H_{\t{int}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{p,\e}^\dag \hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}
    + \hat c_{q,\e}^\dag \hat c_{p,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}^\dag \hat c_{p,\e}
    - \hat c_{q,\e}^\dag \hat c_{q,\g} \hat c_{p,\g}^\dag \hat c_{p,\e}},
\end{align}
where we can use the single-particle pseudo-spin-1/2 Pauli operators
in \eqref{eq:pseudospin} to write
\begin{align}
  H_{\t{int}} = \f{U}{L} \sum_{p,q}
  \sp{\p{\f{\1_q-\sigma_q^z}{2}} \p{\f{\1_p+\sigma_p^z}{2}}
    - \sigma_q^+ \sigma_p^-}
  = \f{U}{L} \sum_{p,q}\f14\p{\1_q \1_p - \v\sigma_q\c\v\sigma_p},
  \label{eq:H_int_pauli}
\end{align}
with
\begin{align}
  \v{\sigma}_p \equiv \p{\sigma_p^x,\sigma_p^y,\sigma_p^z}.
\end{align}
For a fixed total particle number $N$, the identity operators in
\eqref{eq:H_int_pauli} contribute only a global shift in energy, which
allows us to more simply write
\begin{align}
  H_{\t{int}} = - \f{U}{L} \v S\c\v S,
  &&
  \v S \equiv \f12 \sum_p \v{\sigma}_p.
\end{align}
This Hamiltonian has electronic eigenstates $\set{\ket{Sm}}$ with
total (pseudo-)spin $S$ and projection $m$ onto a quantization axis;
the corresponding energies are
$\bk{Sm|H_{\t{int}}|S,m}=-\p{U/L}S\p{S+1}$.  In particular, the
ground-state $S=N/2$ manifold is spanned by the Dicke states
\begin{align}
  \ket{m} \equiv \ket{N/2,m} \propto S_+^{N/2+m} \ket{\g}^{\otimes N},
  &&
  S_+ \equiv \sum_n \sigma_n^+,
  \label{eq:dicke_states}
\end{align}
where $n$ indexes an individual atom,
$\sigma_n^+\equiv\hat c_{n,\e}\hat c_{n,\g}$ is an individual
spin-raising operator, and $N/2+m=0,1,\cdots,N$ is the number of
electronic state excitations in the state $\ket{m}$
\cite{swallows2011suppression}.  In words, the Dicke state $\ket{m}$
is a uniform superposition of all states with $N/2+m$ total electronic
excitations.

@@

\begin{table}[h]
  \centering
  \caption{Example parameters for spin squeezing via one-axis
    twisting.}
  \label{tab:parameters}
  \begin{tabular}{|l|c|l|}
    \hline
    Parameter & Symbol & Value \\ \hline\hline
    Primary lattice depth & $V_0$ & 4 $E_R$ \\
    Transverse lattice depths & $V_T$ & 60 $E_R$ \\
    SOC strength & $\phi$ & $\pi/50$ \\
    Lattice sites & $L$ & 100 \\
    Atom number & $N$ & 50 \\ \hline\hline
    Tunneling rate & $J_0$ & $\approx300\times2\pi$ Hz \\
    On-site interaction strength & $U$ & $\approx1.4\times2\pi$ kHz \\
    Squeezing strength & $\chi$ & $\approx21\times2\pi$ mHz \\
    Optimal squeezing time & $t_{\t{opt}}^{\t{OAT}}$
    & $\approx0.59$ seconds \\ \hline
  \end{tabular}
\end{table}

@@

We can go beyond the $\sim1/N^{2/3}$ one-axis twisting noise floor and
achieve an improved $\sim1/N$ noise floor via so-called two-axis
twisting, which can be achieved with external driving protocols.  By
turning on a clock laser resonant on the electronic transition of the
atoms, we can realize the Hamiltonian
\begin{align}
  H_{\t{drive}} = -\Omega \sum_n s_n^x + H_{\t{OAT}}^z
  = -\Omega S_x + \chi S_z^2,
\end{align}
where the drive strength $\Omega$ may generally depend on time, but
must always be much smaller in magnitude than the collective spin gap
$\eta U$ in order preserve the validity of the effective one-axis
twisting Hamiltonian (see Appendix \ref{sec:squeezing_derivation}).
Following the prescription in ref.~\cite{huang2015twoaxis} we now
modulate the drive as $\Omega\p{t}=\beta\omega\cos\p{\omega t}$ with
$\omega\gg N\chi$, and move into the rotating frame of
$-\Omega\p{t}S_x$.  After a single secular approximation which relies
only on $\omega\gg N\chi$, this procedure results in the effective
Hamiltonian
\begin{align}
  H_{\t{drive,eff}}^{\t{mod}}
  = \f{\chi}{2} \p{\sp{\J_0\p{2\beta}+1} S_z^2
    - \sp{\J_0\p{2\beta}-1} S_y^2},
\end{align}
where $\J_0$ is the zero-order Bessel function of the first kind.
Choosing a modulation index $\beta$ such that $\J_0\p{2\beta}=1/3$ or
$\J_0\p{2\beta}=-1/3$ then respectively results in the two-axis
twisting Hamiltonians
\begin{align}
  H_{\t{TAT}}^{z,x}
  = \f{\chi}{3} \p{2 S_z^2 + S_y^2}
  \simeq \f{\chi}{3} \p{S_z^2 - S_x^2},
  &&
  H_{\t{TAT}}^{y,x}
  = \f{\chi}{3} \p{S_z^2 + 2 S_y^2}
  \simeq \f{\chi}{3}\p{S_y^2 - S_x^2},
  \label{eq:H_TAT_drive}
\end{align}
where $\simeq$ denotes equality up to a global energy shift.  Given
the set of parameters in Table \ref{tab:parameters}, the two-axis
twisting Hamiltonian $H_{\t{TAT}}^{z,x}$ can be achieved with a drive
frequency $\omega=\abs{\eta U\chi}^{1/2}\approx27\times2\pi$ Hz and
modulation index $\beta\approx0.906$, and has an optimal squeezing
time of $t_{\t{opt}}^{\t{TAT}}\approx0.82$ seconds.  The figure of
merit for squeezing is the normalized minimal variance of spin in the
plane orthogonal to the mean spin vector $\bk{\v S}$:
\begin{align}
  \xi^2 \equiv \f{N}{\abs{\bk{\v S}}^2}
  \min_{\hat{\v n}\perp\bk{\v S}} \bk{\p{\v S\c\hat{\v n}}^2}
\end{align}
for $\abs{\hat{\v n}}=1$.  A comparison of this squeezing parameter
through $H_{\t{OAT}}^z$ and $H_{\t{TAT}}^{z,x}$ for the initial state
$\ket{Y}\equiv\bigotimes_n\p{\ket{\g}_n+i\ket{\e}_n}/\sqrt2$ via the
parameters in Table \ref{tab:parameters} is provided in Figure
\ref{fig:squeezing_comparison_TAT}.

\begin{figure}
  \centering \includegraphics{squeezing_comparison_TAT.pdf}
  \caption{Comparison of squeezing through $H_{\t{OAT}}^z$ and
    $H_{\t{TAT}}^{z,x}$ via the parameters in Table
    \ref{tab:parameters} and the continuous two-axis twisting protocol
    with $\omega=27\times2\pi$ Hz and $\beta=0.906$.}
  \label{fig:squeezing_comparison_TAT}
\end{figure}

@@

For small system sizes, we can verify the validity of the effective
one- and two-axis twisting Hamiltonians via direct simulations of a
single-band Fermi-Hubbard (FH) model, using the spin-orbit coupling
Hamiltonian $H_{\t{SOC}}$ and interaction Hamiltonian $H_{\t{int}}$
respectively given in \eqref{eq:H_SOC_start} and
\eqref{eq:H_int_full}.  Figure \ref{fig:squeezing_comparison_FH} shows
a comparison of the squeezing parameter $\xi^2$ after evolution of the
initial state $\ket{Y}$ for $N=8$ particles in a 1-D lattice with
$L=8$ sites (and all remaining parameters as given in Table
\ref{tab:parameters}) after evolution under
\begin{enumerate*}[label=(\roman*)]
\item the one-axis twisting Hamiltonian $H_{\t{OAT}}^z$,
\item the two-axis twisting Hamiltonian $H_{\t{OAT}}^{z,x}$,
\item free evolution of the Fermi-Hubbard model under
  $H_{\t{SOC}}+H_{\t{int}}$, and
\item driven evolution of the Fermi-Hubbard model using the
  continuous-drive protocol in Section \ref{sec:continuous_drive} with
  drive frequency $\omega=\abs{\eta U\chi}^{1/2}$ and modulation index
  $\beta=0.906$.
\end{enumerate*}
These simulation results show essentially perfect agreement between
the Fermi-Hubbard model and the one- and two-axis twisting
Hamiltonians for the given parameters.

\begin{figure}
  \centering
  \includegraphics{squeezing_comparison_FH.pdf}
  \caption{Benchmarking squeezing under the one- and two-axis
    Hamiltonians $H_{\t{OAT}}^z$ and $H_{\t{TAT}}^{z,x}$ against
    direct simulations of the Fermi-Hubbard model with $N=L=8$ and the
    remaining lattice/SOC parameters as given in Table
    \ref{tab:parameters}.}
  \label{fig:squeezing_comparison_FH}
\end{figure}

@@

It may be possible to further enhance the twisting strength $\chi$ by
placing the atoms in an optical cavity, as in
ref.~\cite{hu2017vacuum}.  Both squeezing times and squeezing
strengths might also be further enhanced by using a 2-D lattice,
although more benchmarking is necessary to verify the validity of the
collective spin model in 2-D.  As shown in Figure
\ref{fig:squeezing_comparison_2D},

\begin{figure}
  \centering
  \includegraphics{squeezing_comparison_2D.pdf}
  \caption{Squeezing via one- and two-axis twisting in a 2-D lattice
    with $100\times100$ sites at half filling, with the remaining
    lattice/SOC parameters given in Table \ref{tab:parameters}.}
  \label{fig:squeezing_comparison_2D}
\end{figure}

>>>>

We start with two-level atoms loaded into the ground band of an
optical lattice, and consider interrogating these atoms by a linearly
polarized plane-wave laser tuned to the relevant electronic (atomic)
transition, with wavenumber projection $\phi$ onto the lattice axis.
After performing a gauge transformation which shifts the on-axis
momenta of ground (excited) states by $\phi/2$ ($-\phi/2$), and
eliminating (i.e.~turning off) the interrogation laser, one can arrive
at the Hamiltonian
\begin{align}
  H_{\t{SOC}}
  = \sum_{q,s} E_{qs\phi} \hat c_{qs}^\dag \hat c_{qs}
  \approx -2 J_0\sum_{q,s} \cos\p{q+s\phi/2}
  \hat c_{qs}^\dag \hat c_{qs}
  \label{eq:H_SOC_start}
\end{align}
where $q$ indexes quasi-momentum along the lattice axis,
$s\in\set{\g\leftrightarrow-1,\e\leftrightarrow1}$ labels the
electronic state, $J_0$ is the (positive) ground-band tunneling rate,
$c_{q\sigma}$ is a fermionic annihilation operator, and we work in
units with the lattice spacing $a=1$.  Defining
\begin{align}
  \1_q \equiv \hat c_{q,\g}^\dag \hat c_{q,\g}
  + \hat c_{q,\e}^\dag \hat c_{q,\e},
  &&
  \sigma_q^j \equiv \sum_{\alpha,\beta}
  \hat c_{q\alpha}^\dag \sigma^j_{\alpha\beta} \hat c_{q\beta},
  \label{eq:pseudospin}
\end{align}
for Pauli matrices $\sigma^j$ (i.e.~such that $\sigma_q^j$ is a
pseudo-spin-1/2 Pauli operator on the electronic degrees of freedom)
and expanding the cosine in \eqref{eq:H_SOC_start}, we thus find
\begin{align}
  H_{\t{SOC}}
  = -\sum_q \p{\epsilon_q \1_q + \f12 h_q \sigma_q^z},
  &&
  \epsilon_q \equiv 2J_0 \cos\p{\phi/2} \cos q,
  &&
  h_q \equiv -4J_0 \sin\p{\phi/2} \sin q.
\end{align}
When all dynamics conserve the occupations of all quasi-momenta, the
first term in this Hamiltonian merely contributes a constant energy
shift which we can safely neglect, leaving us with the effective
inhomogeneous-field Hamiltonian
\begin{align}
  H_{\t{SOC}} = -\sum_q h_q s_q^z,
  &&
  s_q^z \equiv \f12 \sigma_q^z
  \label{eq:H_SOC}
\end{align}

@@

At ultracold temperatures inter-atomic interactions are dominated by
$s$-wave collisions, which are captured by the Hamiltonian
\begin{align}
  H_{\t{int}} = G \int \d^3x~
  \hat\psi_\e^\dag\p{x} \hat\psi_\g^\dag\p{x}
  \hat\psi_\g\p{x} \hat\psi_\e\p{x},
  &&
  G \equiv \f{4\pi a_{\e\g^-}}{m_A},
  \label{eq:H_int_full}
\end{align}
where $\hat\psi_\sigma$ is a fermionic field operator for atoms in
electronic state $\sigma\in\set{\g,\e}$, $a_{eg}^-$ is an scattering
length, and $m_A$ is the mass of a single atom.  Considering only
occupation of the ground band in a lattice, we can expand the field
operators as
\begin{align}
  \hat\psi_\sigma\p{x} = \sum_q \phi_q\p{x} \hat c_{q\sigma},
\end{align}
for momentum-space (Bloch) wavefunctions $\phi_q$ and fermionic
annihilation operators $\hat c_{q\sigma}$.  The interaction
Hamiltonian then becomes
\begin{align}
  H_{\t{int}} = G \sum_{p,q,r,s} K^{pq}_{rs}
  \hat c_{r,\e}^\dag \hat c_{s,\g}^\dag \hat c_{q,\g} \hat c_{p,\e},
  &&
  K^{pq}_{rs} \equiv \int \d^3x~
  \phi_r\p{x}^* \phi_s\p{x}^* \phi_q\p{x} \phi_p\p{x}.
\end{align}
In a lattice with $L$ sites centered on positions $x_j$ (i.e.~for
$j=1,\cdots L$ indexing lattice site) and localized (Wannier)
wavefunctions $w_j$, we can expand
$\phi_p\p{x}=L^{-1/2}\sum_je^{-ipx_j}w_j\p{x}$, which implies
\begin{align}
  K^{pq}_{rs} = L^{-2} \sum_{j,k,\ell,m} \int \d^3x~
  \exp\sp{-i\p{px_j+qx_k-rx_\ell-sx_m}}
  w_\ell\p{x}^* w_m\p{x}^* w_k\p{x} w_j\p{x}.
\end{align}
If we now assume that we can neglect integrals without $j=k=\ell=m$,
which is equivalent to neglecting inter-site interactions and
interaction-assisted hopping, then we have
\begin{align}
  K^{pq}_{rs} \approx L^{-2} \sum_j \int \d^3x~
  \exp\sp{-i\p{p+q-r-s}x_j} \abs{w_j\p{x}}^4,
\end{align}
where a stationary phase approximation now forces $p+q-r-s=0$ (mod
$2\pi$ in units with the lattice spacing $a=1$), which is equivalent
to conservation of total momentum.  In terms of the Kronecker delta
$\delta_{p+q,r+s}$ enforcing $p+q=r+s$, it follows that
\begin{align}
  K^{pq}_{rs} \approx \delta_{p+q,r+s} L^{-2}
  \sum_j \int \d^3x~ \abs{w_j\p{x}}^4
  = \delta_{p+q,r+s} L^{-1} \int \d^3x~ \abs{w_0\p{x}}^4,
\end{align}
which motivates the definition
\begin{align}
  U \equiv G \int \d^3x~ \abs{w_0\p{x}}^4,
\end{align}
in order to express the interaction Hamiltonian in the simple form
\begin{align}
  H_{\t{int}} = \f{U}{L} \sum_{p,q,r,s}
  \delta_{p+q,r+s} \hat c_{r,\e}^\dag \hat c_{s,\g}^\dag
  \hat c_{q,\g} \hat c_{p,\e}.
  \label{eq:H_int_uniform}
\end{align}

@@

If the single-particle energy spacings are large compared to the
strength of inter-particle interactions, or equivalently if
$J_0\gtrsim U$, then by the secular approximation we can neglect terms
in $H_{\t{int}}$ which do not conserve the sum of single-particle
energies.  By solving this single-particle energy conservation
condition, one finds that we must have $\set{p,q}\ne\set{r,s}$; that
is, atoms in different clock states can only interact via
\begin{enumerate*}[label=(\roman*)]
\item direct density-density terms, and
\item terms which exchange their momenta.
\end{enumerate*}
The surviving terms in \eqref{eq:H_int_uniform} are thus
\begin{align}
  H_{\t{int}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{p,\e}^\dag \hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}
    + \hat c_{q,\e}^\dag \hat c_{p,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}^\dag \hat c_{p,\e}
    - \hat c_{q,\e}^\dag \hat c_{q,\g} \hat c_{p,\g}^\dag \hat c_{p,\e}},
\end{align}
where we can use the single-particle pseudo-spin-1/2 Pauli operators
in \eqref{eq:pseudospin} to write
\begin{align}
  H_{\t{int}} = \f{U}{L} \sum_{p,q}
  \sp{\p{\f{\1_q-\sigma_q^z}{2}} \p{\f{\1_p+\sigma_p^z}{2}}
    - \sigma_q^+ \sigma_p^-}
  = \f{U}{L} \sum_{p,q}\f14\p{\1_q \1_p - \v\sigma_q\c\v\sigma_p},
  \label{eq:H_int_pauli}
\end{align}
with
\begin{align}
  \v{\sigma}_p \equiv \p{\sigma_p^x,\sigma_p^y,\sigma_p^z}.
\end{align}
For a fixed total particle number $N$, the identity operators in
\eqref{eq:H_int_pauli} contribute only a global shift in energy, which
allows us to more simply write
\begin{align}
  H_{\t{int}} = - \f{U}{L} \v S\c\v S,
  &&
  \v S \equiv \f12 \sum_p \v{\sigma}_p.
\end{align}
This Hamiltonian has electronic eigenstates $\set{\ket{Sm}}$ with
total (pseudo-)spin $S$ and projection $m$ onto a quantization axis;
the corresponding energies are
$\bk{Sm|H_{\t{int}}|S,m}=-\p{U/L}S\p{S+1}$.  In particular, the
ground-state $S=N/2$ manifold is spanned by the Dicke states
\begin{align}
  \ket{m} \equiv \ket{N/2,m} \propto S_+^{N/2+m} \ket{\g}^{\otimes N},
  &&
  S_+ \equiv \sum_n \sigma_n^+,
  \label{eq:dicke_states}
\end{align}
where $n$ indexes an individual atom,
$\sigma_n^+\equiv\hat c_{n,\e}\hat c_{n,\g}$ is an individual
spin-raising operator, and $N/2+m=0,1,\cdots,N$ is the number of
electronic state excitations in the state $\ket{m}$
\cite{swallows2011suppression}.  In words, the Dicke state $\ket{m}$
is a uniform superposition of all states with $N/2+m$ total electronic
excitations.

@@

\begin{table}[h]
  \centering
  \caption{Example parameters for spin squeezing via one-axis
    twisting.}
  \label{tab:parameters}
  \begin{tabular}{|l|c|l|}
    \hline
    Parameter & Symbol & Value \\ \hline\hline
    Primary lattice depth & $V_0$ & 5 $E_R$ \\
    Transverse lattice depths & $V_T$ & 50 $E_R$ \\
    SOC strength & $\phi$ & $\pi/50$ \\
    Lattice sites & $L$ & 100 \\
    Atom number & $N$ & 100 \\ \hline\hline
    Tunneling rate & $J_0$ & $\approx230\times2\pi$ Hz \\
    On-site interaction strength & $U$ & $\approx1.3\times2\pi$ kHz \\
    Twisting strength & $\chi$ & $\approx3.2\times2\pi$ mHz \\
    Optimal OAT squeezing time & $t_{\t{opt}}^{\t{OAT}}$
    & $\approx2.5$ seconds \\ \hline
  \end{tabular}
\end{table}

@@

We can go beyond the $\sim1/N^{2/3}$ one-axis twisting noise floor and
achieve an improved $\sim1/N$ noise floor via so-called two-axis
twisting, which can be achieved with external driving protocols.  By
turning on a clock laser resonant on the electronic transition of the
atoms, we can realize the Hamiltonian
\begin{align}
  H_{\t{drive}} = -\Omega \sum_n s_n^x + H_{\t{OAT}}^z
  = -\Omega S_x + \chi S_z^2,
\end{align}
where the drive strength $\Omega$ may generally depend on time, but
must always be much smaller in magnitude than the collective spin gap
$\eta U$ in order preserve the validity of the effective one-axis
twisting Hamiltonian (see Appendix \ref{sec:squeezing_derivation}).
Following the prescription in ref.~\cite{huang2015twoaxis} we now
modulate the drive as $\Omega\p{t}=\beta\omega\cos\p{\omega t}$ with
$\omega\gg N\chi$, and move into the rotating frame of
$-\Omega\p{t}S_x$.  After a single secular approximation which relies
only on $\omega\gg N\chi$, this procedure results in the effective
Hamiltonian
\begin{align}
  H_{\t{drive,eff}}^{\t{mod}}
  = \f{\chi}{2} \p{\sp{\J_0\p{2\beta}+1} S_z^2
    - \sp{\J_0\p{2\beta}-1} S_y^2},
\end{align}
where $\J_0$ is the zero-order Bessel function of the first kind.
Choosing a modulation index $\beta$ such that $\J_0\p{2\beta}=1/3$ or
$\J_0\p{2\beta}=-1/3$ then respectively results in the two-axis
twisting Hamiltonians
\begin{align}
  H_{\t{TAT}}^{z,x}
  = \f{\chi}{3} \p{2 S_z^2 + S_y^2}
  \simeq \f{\chi}{3} \p{S_z^2 - S_x^2},
  &&
  H_{\t{TAT}}^{y,x}
  = \f{\chi}{3} \p{S_z^2 + 2 S_y^2}
  \simeq \f{\chi}{3}\p{S_y^2 - S_x^2},
  \label{eq:H_TAT_drive}
\end{align}
where $\simeq$ denotes equality up to a global energy shift.  Given
the set of parameters in Table \ref{tab:parameters}, the two-axis
twisting Hamiltonian $H_{\t{TAT}}^{z,x}$ can be achieved with a drive
frequency $\omega=\abs{\eta U\chi}^{1/2}\approx20\times2\pi$ Hz and
modulation index $\beta\approx0.906$, and has an optimal squeezing
time of $t_{\t{opt}}^{\t{TAT}}\approx3.2$ seconds.  The figure of
merit for squeezing is the normalized minimal variance of spin in the
plane orthogonal to the mean spin vector $\bk{\v S}$:
\begin{align}
  \xi^2 \equiv \f{N}{\abs{\bk{\v S}}^2}
  \min_{\hat{\v n}\perp\bk{\v S}} \bk{\p{\v S\c\hat{\v n}}^2}
\end{align}
for $\abs{\hat{\v n}}=1$.  A comparison of this squeezing parameter
through $H_{\t{OAT}}^z$ and $H_{\t{TAT}}^{z,x}$ for the initial state
$\ket{Y}\equiv\bigotimes_n\p{\ket{\g}_n+i\ket{\e}_n}/\sqrt2$ via the
parameters in Table \ref{tab:parameters} is provided in Figure
\ref{fig:squeezing_comparison_TAT}.

\begin{figure}
  \centering \includegraphics{squeezing_comparison_TAT.pdf}
  \caption{Comparison of squeezing through $H_{\t{OAT}}^z$ and
    $H_{\t{TAT}}^{z,x}$ via the parameters in Table
    \ref{tab:parameters} and the continuous two-axis twisting protocol
    with $\omega\approx\abs{\eta U\chi}^{1/2}$ and $\beta=0.906$.}
  \label{fig:squeezing_comparison_TAT}
\end{figure}

@@

For small system sizes, we can verify the validity of the effective
one- and two-axis twisting Hamiltonians via direct simulations of a
single-band Fermi-Hubbard (FH) model, using the spin-orbit coupling
Hamiltonian $H_{\t{SOC}}$ and interaction Hamiltonian $H_{\t{int}}$
respectively given in \eqref{eq:H_SOC_start} and
\eqref{eq:H_int_full}.  Figure \ref{fig:squeezing_comparison_FH} shows
a comparison of the squeezing parameter $\xi^2$ after evolution of the
initial state $\ket{Y}$ for $N=8$ particles in a 1-D lattice with
$L=8$ sites (and remaining system parameters as given in the top half
of Table \ref{tab:parameters}) after evolution under
\begin{enumerate*}[label=(\roman*)]
\item the one-axis twisting Hamiltonian $H_{\t{OAT}}^z$,
\item the two-axis twisting Hamiltonian $H_{\t{OAT}}^{z,x}$,
\item free evolution of the Fermi-Hubbard model under
  $H_{\t{SOC}}+H_{\t{int}}$, and
\item driven evolution of the Fermi-Hubbard model using the
  continuous-drive protocol in Section \ref{sec:continuous_drive} with
  drive frequency $\omega=\abs{\eta U\chi}^{1/2}$ and modulation index
  $\beta=0.906$.
\end{enumerate*}
These simulation results show essentially perfect agreement between
the Fermi-Hubbard model and the one- and two-axis twisting
Hamiltonians for the given parameters.

\begin{figure}
  \centering
  \includegraphics{squeezing_comparison_FH.pdf}
  \caption{Benchmarking squeezing under the one- and two-axis
    Hamiltonians $H_{\t{OAT}}^z$ and $H_{\t{TAT}}^{z,x}$ against
    direct simulations of the Fermi-Hubbard model with $N=L=8$ and the
    remaining system parameters as given in the top half of Table
    \ref{tab:parameters}.}
  \label{fig:squeezing_comparison_FH}
\end{figure}

@@

It may be possible to further enhance the twisting strength $\chi$ by
placing the atoms in an optical cavity, as in
ref.~\cite{hu2017vacuum}.  Both squeezing times and twisting strengths
might also be further enhanced by using a 2-D lattice, although more
benchmarking is necessary to verify the validity of the collective
spin model in 2-D.  As shown in Figure
\ref{fig:squeezing_comparison_2D},

\begin{figure}
  \centering
  \includegraphics{squeezing_comparison_2D.pdf}
  \caption{Squeezing via one- and two-axis twisting in a 2-D lattice
    with $100\times100$ sites at unit filling, with the remaining
    lattice/SOC parameters given in Table \ref{tab:parameters}.}
  \label{fig:squeezing_comparison_2D}
\end{figure}
